---
title: "Weather Event's Impact on Health and Ecnomics"
author: "Chuk Yong"
date: "19 July 2016"
output:
  html_fragment: default
  md_document:
    variant: markdown_github
    pandoc_args: "--no-wrap"
---

## Impact of Severe Weather Events on Public Health and Economy in the United States of America

Reproducible Research Peer Assessment
Author: Chuk Yong
Date: 20 July 2016

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to identify the following:

1. Which types of weather events are most harmful to population health?
2. Which types of weather events have the greatest economic impact?

### Data Processing

Load and Read in the storm data set.
```{r read, cache=TRUE}
setwd("~/Desktop/Rstudio/RepData_PeerAssessment2")
Data <- read.csv("repdata-data-StormData.csv")
```

Load the neccessary libraries
```{r echo=FALSE}
library(dplyr)
library(ggplot2)
```

Extract data for the top 20 weather events that caused the most fatality
```{r fatal, cache=TRUE}
FataData <- Data %>% group_by(EVTYPE) %>% summarise(FATAL=sum(FATALITIES)+sum(INJURIES)) %>% arrange(desc(FATAL))
FataDataClean <- FataData[1:20,]
FataDataClean$EVTYPE <- factor(FataDataClean$EVTYPE, levels = FataDataClean$EVTYPE[order(FataDataClean$FATAL)])
```

Extract data for the top 20 weather events that caused the most economic damage

```{r}
EconData <- Data %>%
  filter(PROPDMG>0 | Data$CROPDMG>0) %>%
  select(EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP) %>%
  mutate(PROPDMG = PROPDMG*(ifelse(PROPDMGEXP == "K",1000, ifelse(PROPDMGEXP == "M",1000000,ifelse(PROPDMGEXP == "B",1000000000,1))))) %>%
  mutate(CROPDMG = CROPDMG*(ifelse(CROPDMGEXP == "K",1000, ifelse(CROPDMGEXP == "M",1000000,ifelse(CROPDMGEXP == "B",1000000000,1))))) %>%
  mutate(TOTDMG = PROPDMG+CROPDMG)

EconDataSum <- EconData %>% group_by(EVTYPE) %>% 
  summarise(TOTECONDMG = sum(TOTDMG)) %>%
  arrange(desc(TOTECONDMG)) %>%
  slice(1:20)
```

## Results of the study

```{r plot_fatal, echo=FALSE, message=FALSE}
plot <- ggplot(FataDataClean, aes(x=FataDataClean$EVTYPE, y=FataDataClean$FATAL)) +
  geom_bar(stat = "identity") + 
  # Flip the x-y so that the "Type" is easier to read
  coord_flip() +
  xlab("Type of Event") + ylab("Number of Fatality") + ggtitle("Top 20 Fatal Weather Events in the United States")
plot
```

From the graph, the most fatal weather event was tornado. The fatality rate caused by tornado far exceeded the other weather events.

----------------------------------------------------------
```{r plot_economic_damage, echo=FALSE, message=FALSE}
EconDataSum$EVTYPE <- factor(EconDataSum$EVTYPE, levels = EconDataSum$EVTYPE[order(EconDataSum$TOTECONDMG)])
plot <- ggplot(EconDataSum, aes(x=EconDataSum$EVTYPE, y=EconDataSum$TOTECONDMG)) +
  geom_bar(stat = "identity") +
  # Flip the x-y so that the "Type" is easier to read
  coord_flip() + 
  xlab("") + ylab("Economic Cost ") + ggtitle("Top 20 Most Costly Weather Events (USA)")
plot

```

The weather event that caused the most economic damage is flood, followed by hurricane/typhoon, tornado and storm surge.
